---
title: "2. Loan Data"
author: "Huy Nguyen"
date: "2023-07-05"
output: html_document
---

```{r}
load("Cleaned Data.RData")
library(tidyverse)
```

## Months Employed
### by Age Group
```{r}
ggplot(dat, aes(ageGroup, monthsEmployed, fill = ageGroup))+
  geom_bar(stat = "summary", fun = mean)+
  theme_classic()+
  labs(title = "Mean Months Employed by Age Group")
#Consistent
```
### by Gender
```{r}
ggplot(dat, aes(gender, monthsEmployed, fill = gender))+
  geom_bar(stat = "summary", fun = mean)+
  theme_classic()+
  labs(title = "Mean Months Employed by gender")
```

### by Credit Risk Category
```{r}
ggplot(dat, aes(creditRisk, monthsEmployed, fill = creditRisk))+
  geom_bar(stat = "summary", fun = mean)+
  theme_classic()+
  labs(title = "Mean Months Employed by Credit Risk")
#Check if difference is significant
```
### by Marital Status
```{r}
ggplot(dat, aes(maritalStat, monthsEmployed, fill = maritalStat))+
  geom_bar(stat = "summary", fun = mean)+
  theme_classic()+
  labs(title = "Mean Months Employed by Marital Status")
```
### by Job Type
```{r}
ggplot(dat, aes(job, monthsEmployed, fill = job))+
  geom_bar(stat = "summary", fun = mean)+
  theme_classic()+
  labs(title = "Mean Months Employed by Job Type")
#Consistent
```

### by monthsEmployed
```{r}
ggplot(dat, aes(yearsCustomer, monthsEmployed, fill = yearsCustomer))+
  geom_bar(stat = "summary", fun = mean)+
  theme_classic()+
  labs(title = "Mean Months Employed by Year Customer")
```


### by Housing
```{r}
ggplot(dat, aes(housing, monthsEmployed, fill = housing))+
  geom_bar(stat = "summary", fun = mean)+
  theme_classic()+
  labs(title = "Mean Months Employed by Housing Status")
```

## Months Customer

### by Age Group
```{r}
ggplot(dat, aes(ageGroup, monthsCustomer, fill = ageGroup))+
  geom_boxplot()+
  theme_classic()
```

### by Gender
```{r}
ggplot(dat, aes(gender, monthsCustomer, fill = gender))+
  geom_boxplot()+
  theme_classic()
```

### by Credit Risk Category
```{r}
ggplot(dat, aes(creditRisk, monthsCustomer, fill = creditRisk))+
  geom_boxplot()+
  theme_classic()
```

### by Marital Status
```{r}
ggplot(dat, aes(maritalStat, monthsCustomer, fill = maritalStat))+
  geom_boxplot()+
  theme_classic()
```

### by Job Type
```{r}
ggplot(dat, aes(job, monthsCustomer, fill = job))+
  geom_boxplot()+
  theme_classic()
```

### by Loan Type
```{r}
ggplot(dat, aes(fct_reorder(loanPurpose, age, .desc = T), age, fill = loanPurpose))+
  geom_violin()+
  geom_boxplot(alpha = .5)+
  stat_summary(fun = mean, geom= "point", shape = 21, size = 3, color = "black", fill = "red")+
  theme_classic()

ggplot(dat, aes(fct_reorder(loanPurpose, monthsCustomer, .desc = T), monthsCustomer, fill = loanPurpose))+
  geom_violin()+
  geom_boxplot(alpha = .5)+
  stat_summary(fun = mean, geom= "point", shape = 21, size = 3, color = "black", fill = "red")+
  theme_classic()
```

```{r}
cstLfSpnCr <- dat %>% filter(loanPurpose %in% c("New Car", "Used Car"))

cstLfSpnCr$loanPurpose <- cstLfSpnCr$loanPurpose %>% as.character() %>% as.factor()

paste("Skew")
tapply(cstLfSpnCr$monthsCustomer, cstLfSpnCr$loanPurpose, psych::skew)
paste("Mean")
tapply(cstLfSpnCr$monthsCustomer, cstLfSpnCr$loanPurpose, mean)
wilcox.test(data = cstLfSpnCr, monthsCustomer ~ loanPurpose)
```

## Bank Account

```{r}
library(scales)
ggplot(dat, aes(checking, savings, color = creditRisk))+
  geom_point()+
  geom_smooth(method = "lm", linetype = "dashed")+
  facet_wrap(.~creditRisk)+
  theme_classic()+
  scale_y_continuous(breaks = pretty_breaks(n = 9))+
  scale_x_continuous(breaks = pretty_breaks(n = 8))
```

### Filtering extreme case out
```{r}
money <- dat %>% filter(checking != 0 & savings != 0)
ggplot(money, aes(checking, savings, color = creditRisk))+
  geom_point()+
  geom_smooth(method = "lm", linetype = "dashed")+
  facet_wrap(.~creditRisk)+
  theme_classic()+
  scale_y_continuous(breaks = pretty_breaks(n = 9))+
  scale_x_continuous(breaks = pretty_breaks(n = 8))
```

## Credit Risk
Use proportion alongside raw number

```{r}
ggplot(dat, aes(creditRisk, fill = creditRisk))+
  geom_bar()+
  theme_classic()
#Almost equal distribution
```

## by Accounts
```{r}
dat <- dat %>% mutate(
  sumAccount = checking + savings
  )

money2 <- dat %>% select(creditRisk, checking, savings, sumAccount) %>% pivot_longer(
  cols = 2:4,
  names_to = "type",
  values_to = "numVal"
)

ggplot(money2, aes(creditRisk, numVal, fill = type))+
  geom_bar(stat = "summary", fun = mean, position = "dodge")+
  theme_classic()+
  labs(title = "mean")

ggplot(money2, aes(creditRisk, numVal, fill = type))+
  geom_bar(stat = "summary", fun = median, position = "dodge")+
  theme_classic()+
  labs(title = "median") #Acknowledge skew but median of checking is 0 for both group so can't compare
```

T.Test
```{r}
money2.1 <- dat %>% 
  select(creditRisk, checking, savings, sumAccount)

paste("Mann-Whitney U Test, testing mean difference")
wilcox.test(data = money2.1, checking ~ creditRisk)
wilcox.test(data = money2.1, savings ~ creditRisk)
wilcox.test(data = money2.1, sumAccount ~ creditRisk)
#Fail to reject NULL hypothesis. These differences only exist in this sample. Also consistent with logistic regression

library(coin)
paste("Median Test, testing median difference")
median_test(data = money2.1, checking ~ creditRisk)
median_test(data = money2.1, savings ~ creditRisk)
median_test(data = money2.1, sumAccount ~ creditRisk)
#Fail to reject NULL hypothesis. These differences only exist in this sample. Also consistent with logistic regression
```

### by Customer Lifetime
```{r, fig.height=5, fig.width=9}
clCar <- ggplot(data = cstLfSpnCr, aes(monthsCustomer, fill = creditRisk))+
  geom_boxplot()+
  facet_grid(loanPurpose~.)+
  theme_classic()+
  theme(legend.position = "none")+
  scale_x_continuous(name = "Customer Lifetime (months)")

export::graph2png(clCar, "CustomerLifetime.png")
```

```{r}
paste("skew")
tapply(cstLfSpnCr$monthsCustomer[cstLfSpnCr$loanPurpose == "New Car"], cstLfSpnCr$creditRisk[cstLfSpnCr$loanPurpose == "New Car"], psych::skew)
tapply(cstLfSpnCr$monthsCustomer[cstLfSpnCr$loanPurpose == "Used Car"], cstLfSpnCr$creditRisk[cstLfSpnCr$loanPurpose == "Used Car"], psych::skew)
paste("mean")
tapply(cstLfSpnCr$monthsCustomer[cstLfSpnCr$loanPurpose == "New Car"], cstLfSpnCr$creditRisk[cstLfSpnCr$loanPurpose == "New Car"], mean)
tapply(cstLfSpnCr$monthsCustomer[cstLfSpnCr$loanPurpose == "Used Car"], cstLfSpnCr$creditRisk[cstLfSpnCr$loanPurpose == "Used Car"], mean)
```

```{r}
wilcox.test(data = cstLfSpnCr, cstLfSpnCr$monthsCustomer[cstLfSpnCr$loanPurpose == "New Car"] ~ cstLfSpnCr$creditRisk[cstLfSpnCr$loanPurpose == "New Car"])

var.test(data = cstLfSpnCr, cstLfSpnCr$monthsCustomer[cstLfSpnCr$loanPurpose == "Used Car"] ~ cstLfSpnCr$creditRisk[cstLfSpnCr$loanPurpose == "Used Car"])
t.test(data = cstLfSpnCr, cstLfSpnCr$monthsCustomer[cstLfSpnCr$loanPurpose == "Used Car"] ~ cstLfSpnCr$creditRisk[cstLfSpnCr$loanPurpose == "Used Car"], var.equal = F)
```

### by martial Stat
```{r, fig.width=9, fig.height=5}
maritalStat.Prop <- dat %>% 
  group_by(maritalStat, gender) %>%
  summarise(
    highRisk = length(creditRisk[creditRisk == "High"]) / length(creditRisk) * -1,
    lowRisk =  length(creditRisk[creditRisk == "Low"]) / length(creditRisk)
  ) %>%
  pivot_longer(
    cols = c("highRisk", "lowRisk"),
    names_to = "creditRisk",
    values_to = "proportion"
  ) %>%
  mutate(
    valPerc = scales::percent(proportion)
  )

mrtlProp <- ggplot(maritalStat.Prop, aes(proportion, maritalStat, fill = creditRisk, label = valPerc))+
  geom_bar(stat = "identity", reserve = "single")+
  theme_classic()+
  facet_grid(gender~., scales = "free", space = "free")+
  scale_y_discrete(limits = rev, name = "Marital Status")+
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"),
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank())+
  geom_text(hjust = -.1)+
  labs(title = "Marital Percentage by Credit Risk")
#Divorced decree
#Divorced former spouse might attempt to drag both credit scores down via large purchases (without payment) on joint account post-divorce
#Kids might be a factor -> unsure
#Socioeconomic status might be a better predictor
#FIXME: WHAT WOULD BE THE RECOMMENDATION? IS IT POSSIBLE TO APPLY ETHICS IN THIS CASE
#Straightforward would be tightened awareness toward divorcees.
#Are married men more likely to apply for 'joined' loan than
```

### by age(Group)
```{r, fig.width=12, fig.height=5}
age.Prop <- dat %>% 
  group_by(ageGroup) %>%
  summarise(
    highRisk = length(creditRisk[creditRisk == "High"]) / length(creditRisk) * -1,
    lowRisk =  length(creditRisk[creditRisk == "Low"]) / length(creditRisk)
  ) %>%
  pivot_longer(
    cols = c("highRisk", "lowRisk"),
    names_to = "creditRisk",
    values_to = "proportion"
  ) %>%
  mutate(
    valPerc = scales::percent(proportion)
  )

ageProp <- ggplot(age.Prop, aes(proportion, ageGroup, fill = creditRisk, label = valPerc))+
  geom_col()+
  theme_classic()+
  scale_y_discrete(limits = rev, name = "Age Group")+
  geom_text(hjust = -.1)+
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"),
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank())+
  labs(title = "Age Percentage by Credit Risk")
#Young-Adult has highest credit risk, adults about the same, old tend to have lower credit risks
```

### by Housing
```{r, fig.width=13, fig.height=5}
housing.Prop <- dat %>% 
  group_by(housing) %>%
  summarise(
    highRisk = length(creditRisk[creditRisk == "High"]) / length(creditRisk) * -1,
    lowRisk =  length(creditRisk[creditRisk == "Low"]) / length(creditRisk)
  ) %>%
  pivot_longer(
    cols = c("highRisk", "lowRisk"),
    names_to = "creditRisk",
    values_to = "proportion"
  ) %>%
  mutate(
    valPerc = scales::percent(proportion)
  )

hs.Prop <- ggplot(housing.Prop, aes(proportion, housing, fill = creditRisk, label = valPerc))+
  geom_col()+
  theme_classic()+
  scale_y_discrete(limits = rev, name = "Housing")+
  geom_text(hjust = -.1)+
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"),
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank())+
  labs(title = "Housing Percentage by Credit Risk")
#House owners are more financially stable
```

### by Job
```{r}
job.Prop <- dat %>% 
  group_by(job) %>%
  summarise(
    highRisk = length(creditRisk[creditRisk == "High"]) / length(creditRisk) * -1,
    lowRisk =  length(creditRisk[creditRisk == "Low"]) / length(creditRisk)
  ) %>%
  pivot_longer(
    cols = c("highRisk", "lowRisk"),
    names_to = "creditRisk",
    values_to = "proportion"
  ) %>%
  mutate(
    valPerc = scales::percent(proportion)
  )

ggplot(job.Prop, aes(proportion, job, fill = creditRisk, label = valPerc))+
  geom_col()+
  theme_classic()+
  scale_y_discrete(limits = rev)+
  geom_text(hjust = -.1)
#About the same across the board
```

### by monthsCustomer
```{r, fig.width=12, fig.height=5}
yCust.Prop <- dat %>% 
  group_by(yearsCustomer) %>%
  summarise(
    highRisk = length(creditRisk[creditRisk == "High"]) / length(creditRisk) * -1,
    lowRisk =  length(creditRisk[creditRisk == "Low"]) / length(creditRisk)
  ) %>%
  pivot_longer(
    cols = c("highRisk", "lowRisk"),
    names_to = "creditRisk",
    values_to = "proportion"
  ) %>%
  mutate(
    valPerc = scales::percent(proportion)
  )

cstLF <- ggplot(yCust.Prop, aes(proportion, yearsCustomer, fill = creditRisk, label = valPerc))+
  geom_col()+
  theme_classic()+
  scale_y_discrete(limits = rev, name = "Years")+
  geom_text(hjust = -.1)+
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"),
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank())+
  labs(title = "Customer Lifetime (Years) Percentage by Credit Risk")
```

```{r}
library(patchwork)
mrtlProp + ageProp + hs.Prop + cstLF
```


### by yearsEmployed
```{r}
yEmp.Prop <- dat %>% 
  group_by(yearsEmployed) %>%
  summarise(
    highRisk = length(creditRisk[creditRisk == "High"]) / length(creditRisk) * -1,
    lowRisk =  length(creditRisk[creditRisk == "Low"]) / length(creditRisk)
  ) %>%
  pivot_longer(
    cols = c("highRisk", "lowRisk"),
    names_to = "creditRisk",
    values_to = "proportion"
  ) %>%
  mutate(
    valPerc = scales::percent(proportion)
  )

ggplot(yEmp.Prop, aes(proportion, yearsEmployed, fill = creditRisk, label = valPerc))+
  geom_col()+
  theme_classic()+
  scale_y_discrete(limits = rev)+
  geom_text(hjust = -.1)
#Pretty consistent surprisingly
```

### by Loan Type
```{r}
lnTyp.Prop <- dat %>% 
  group_by(loanPurpose) %>%
  summarise(
    highRisk = length(creditRisk[creditRisk == "High"]) / length(creditRisk) * -1,
    lowRisk =  length(creditRisk[creditRisk == "Low"]) / length(creditRisk)
  ) %>%
  pivot_longer(
    cols = c("highRisk", "lowRisk"),
    names_to = "creditRisk",
    values_to = "proportion"
  ) %>%
  mutate(
    valPerc = scales::percent(proportion)
  )

ggplot(lnTyp.Prop, aes(proportion, fct_reorder(loanPurpose, proportion, .desc = T), fill = creditRisk, label = valPerc))+
  geom_col()+
  theme_classic()+
  geom_text(hjust = -.1)+
  scale_y_discrete(name = "Loan Purpose")+
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"),
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank())+
  labs(title = "Loan Purpose Percentage by Credit Risk")
```


